 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-07-09 paper.Rmd
reknit

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pariwise distances (inner products) in $Y$ 
approximate the pairwise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as nonmetric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbor graph
to estimate the pairwise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such asdiffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pairwise distance matrix and k-nearest neighbors graph of $X$. 
The algorithm denoises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

@@

We propose using tours as part of an analyst's workflow when using embedding
methods like t-SNE. We have made extensive use of ensemble graphics, that is
aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
As we will see in the case studies, this allows analysts to quickly compare 
views fromembedding methods and allows them to see how the embedding method 
alters theglobal structure of their data. Using ensembles allows the use of 
interaction techniques, that allow analysts to perform cluster orientation 
tasks vialinking multiple views. This approach allows our interface, to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

A common analysis task in single cell studies is performing clustering to
identify groupings  of cells with similar expression profiles. Analysts
in this area generally use non linear DR methods for verification and identifaction
of clusters and developmental trajectories (i.e., case study 1). For clustering,
workflows the primary task is verify the existence of clusters and then begin to 
identify the clusters as cell types using the expression of "known" 
marker genes. Here a 'faithful' a embedding should ideally preserve the topology
of the data; cells that correspond to a cell type should lie in the same
neighbourhood in high-dimensional space. In this case study we use our 
spatial and linked brushing approaches to look at neighbourhood preservation
(which can be seen via the tour) and to look at the affect of applying different
feature selection techniques and distance metrics.

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitocondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighours graph (with $k = 10$) and 
used Louvain clustering to generate clusters. To check and verify the 
clustering we construct a liminal view. Because most off the variablity in
expression is explained by the first 5 PCs, we tour those. The t-SNE view
is computed from all PCs

>>>>

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pairwise distances (inner products) in $Y$ 
approximate the pairwise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pairwise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such asdiffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pairwise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

@@

We propose using tours as part of an analyst's workflow when using embedding
methods like t-SNE. We have made extensive use of ensemble graphics, that is
aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
As we will see in the case studies, this allows analysts to quickly compare 
views from embedding methods and allows them to see how the embedding method 
alters the global structure of their data. Using ensembles allows the use of 
interaction techniques, that allow analysts to perform cluster orientation 
tasks via linking multiple views. This approach allows our interface, to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

A common analysis task in single cell studies is performing clustering to
identify groupings  of cells with similar expression profiles. Analysts
in this area generally use non linear DR methods for verification and identification
of clusters and developmental trajectories (i.e., case study 1). For clustering,
workflows the primary task is verify the existence of clusters and then begin to 
identify the clusters as cell types using the expression of "known" 
marker genes. Here a 'faithful' a embedding should ideally preserve the topology
of the data; cells that correspond to a cell type should lie in the same
neighbourhood in high-dimensional space. In this case study we use our 
spatial and linked brushing approaches to look at neighbourhood preservation
(which can be seen via the tour) and to look at the affect of applying different
feature selection techniques and distance metrics.

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters. To check and verify the 
clustering we construct a liminal view. Because most off the variability in
expression is explained by the first 5 PCs, we tour those. The t-SNE view
is computed from all PCs
